---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stochvolTMB

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Build Status](https://travis-ci.org/JensWahl/stochvolTMB.svg?branch=master)](https://travis-ci.org/JensWahl/stochvolTMB)
[![License:
MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

The goal of stochvolTMB is to ...

## Installation

You can install stochvolTMB from github by running 

``` r
# install.packages("devtools")
devtools::install_github("JensWahl/stochvolTMB")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JensWahl/stochvolTMB")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(stochvolTMB)
# load data
data("spy")

# estimate parameters 
opt <- estimate_parameters(spy$log_return, model = "gaussian", silent = TRUE)

# get parameter estimates with standard error
estimates <- summary(opt)
head(estimates, 10)

## plot estimated volatility with 95 % confidence interval
plot(opt, include_ci = TRUE)
```

By running `demo()` you start a shiny application where you can visually inspect the effect of choosing  different models and parameter settings. 
