---
title: "stochvolTMB: Likelihood estimation of stochastic volatility"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{stochvolTMB: Likelihood estimation of stochastic volatility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(stochvolTMB)
```

# The stochastic volatility model

The stochastic volatility model, introduced by @Taylor_SV_1982,  is defined by
\begin{equation}
    \begin{aligned}
        y_t &= \sigma_y e^{h_t/2} \epsilon_t, \quad t = 1, \dots, T, \\
        h_{t+1} &= \phi h_{t} + \sigma_h \eta_t, \quad t = 1, \dots, T-1,
    \end{aligned}
\end{equation}
where $y_t$ is the observed log returns, $h_t$ is the logarithm of the variance on day $t$, $\eta_t \stackrel{\text{iid}}{\sim} \mathcal{N}(0,1)$ and $\epsilon_t \sim F$, where $F$ is specified below. To ensure stationarity for $h_t$, we assume $|\phi| < 1$. It can be shown that the unconditional distribution of $h_t$ is $\mathcal{N}(0,\sigma_h^2/(1 - \phi^2))$, and we assume $h_1 \sim \mathcal{N}(0,\sigma_h^2/(1-\phi^2))$. A interpretation of the latent process $\{h_t\}$ is that is represents the random and uneven flow of new information into the marked. For different time points, the variance will be dependent of this unobserved ``flow'' of information, i.e. conditioning on $h_t$, $\mathrm{Var}(y_t | h_t) = \sigma_x^2 e^{h_t}$.

The original SV @Taylor_SV_1982 assumed normally distributed innovations, but this is usually to strong of an assumption. Financial returns are usually heavy-tailed, might be asymmetric and the volatility can be correlated with the returns. The latter is called the *leverage effect*, where there is a negative correlation between a change in price and the volatilty, meaning that a drop in price tend to lead to an increase in volatilty. To take this features of financial time series into account `stochvolTMB` has implemented the following four distribution for the inovation: 

* Gaussian: $\epsilon_t \sim \mathcal{N}(0,1)$
* t: $\epsilon_t \sim t_\nu(0,1)$, where $\nu$ is the degrees of freedom. 
* Skew-Gaussian: $\epsilon_t \sim \mathcal{SN}(0,1, \alpha)$, with zero mean, unit variance and skrewness parameter $\alpha$. 
* Leverage: $\epsilon_t \sim \mathcal{N}(0,1)$ and $\mathrm{Cor}(\epsilon_t, \eta_t) = \rho$

# Usage

The main functions of `stochvolTMB` are: 

----------------------------- ------------------------------------------------------
Function Name                 Description
----------------------------- ------------------------------------------------------
`estimate_parameters`         Estimate parameters of a stochastic volatility model.

`sim_sv`                      Simulate data from a stochastic volatility model.

`plot-stochvolTMB`            Plot estimated volatility (with confidence intervalls).

`summary.stochvolTMB`         Extract parameter estimates and volatility with uncertainty.
----------------------------- --------------------------------------------------

  Estimating parameters in a SV model is easy. As an example we investigate the daily log-returns of the S&P500 from 2005 to 2018. We fit all four distributions: 
  
  
```{r warning=FALSE}
library(stochvolTMB)
data(spy)

guassian = estimate_parameters(spy$log_return, model = "gaussian", silent = TRUE)
t_dist = estimate_parameters(spy$log_return, model = "t", silent = TRUE)
skew_gaussian = estimate_parameters(spy$log_return, model = "skew_gaussian", silent = TRUE)
leverage = estimate_parameters(spy$log_return, model = "leverage", silent = TRUE)

```

We can investigate the estimate for the degrees of freedom (`df`) to see if the returns are heavy-tailed

```{r}
summary(t_dist, report = "fixed")
```

Clearly the returns are more heavy tailed than Gaussian, even when controlling for the stochastic volatility. We can also check for asymetric returns

```{r}
summary(skew_gaussian, report = "fixed")
```

and leverage (`rho`)

```{r}
summary(leverage, report = "transformed")
```
There is clear evidence for both asymetric returns and a negative correlation (of 0.74!) between log-returns and the volatility. To find the model that fits the data best, we can compare the [AIC](https://en.wikipedia.org/wiki/Akaike_information_criterion) of out models and pick the smallest. 

```{r}

AIC(gaussian, 
    t_dist, 
    skew_gaussian, 
    leverage)

```


# References